State of the Practice for Medical Imaging Software Based on Open Source Repositories

### 2024-10-29 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Edits up to Section 6 (inclusive)

Our investigation is based on the following four research questions:
\begin{enumerate}
\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
{open source} software projects exist? (Section~\ref{ch_results})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Based on
quantitative measurements of each project's development practices, which
projects follow best practices? (Section~\ref{ch_results})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
similar are the top projects identified in \rqref{RQ_HighestQuality}
to the most popular projects as viewed by the community?
(Section~\ref{Sec_VsCommunityRanking})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
do MI projects compare to general research software with respect to the
artifacts (documents, scripts and code) present in their repositories?
(Section~\ref{Sec_CompareArtifacts})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
How do MI projects compare to research software in general with respect to
the use of tools (Section~\ref{Sec_CompareTools}) for:
\begin{enumerate} 
	\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
	\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
\end{enumerate}
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
How do MI projects compare to research software in general with respect to
principles, processes, and methodologies used?
(Section~\ref{Sec_CompareMethodologies})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
the pain points for developers working on MI software projects?
(Section~\ref{painpoints})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
do the pain points of developers from MI compare to the pain points
for research software in general? (Section~\ref{painpoints})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For MI
developers what specific best practices are taken to address the pain points
and software quality concerns? (Section~\ref{painpoints})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
What research software development practice could potentially address the
pain point concerns identified in \rqref{RQ_PainPoints}?
(Section~\ref{ch_recommendations})
\end{enumerate}

@@

\section{Results} \label{ch_results}

In this section we answer \rqref{RQ_WhatProjects} and \rqref{RQ_HighestQuality}.

\subsection{In-Scope Open-Source MI Software} \label{SecWhatProjects}

@@

\subsection{Which Projects Follow Best Practices?}

@@

[Summary description list]

\subsection{Installability} \label{sec_result_installability}

We found
installation instructions for 16 projects, but two did not need them
(\textit{BioImage Suite Web} and \textit{Slice:Drop})
as they are web applications. 10 of the projects required extra
dependencies: Five depend on a specific browser; \textit{dwv}, \textit{OHIF
Viewer}, and \textit{GATE} needs extra libraries to build; \textit{ImageJ} and
\textit{Fiji} need an unzip tool; \textit{MatrixUser} needs Matlab;
\textit{DICOM Viewer} needs a Nextcloud platform.

The scores are based on the ease of following the installation instructions,
and automated installation and uninstallation process. There were no issues
for all but the bottom four, just various degrees of ease and automation.
\textit{GATE}, \textit{dwv}, and \textit{DICOM Viewer} showed severe
installation problems. We were not able to install them, even after a reasonable
amount of time (2 hours).  For \textit{dwv} and \textit{GATE} we failed to build
from the source code, but we were able to proceed with measuring other qualities
using a deployed on-line version for \textit{dwv}, and a virtual machine version for
\textit{GATE}. For \textit{DICOM Viewer} we could not install the NextCloud
dependency, and thus could not measure reliability nor robustness for it.

\textit{MatrixUser} depends on Matlab, whose installation is not easy.
For users who already have Matlab, this score should be higher.

\subsection{Correctness \& Verifiability} \label{sec_result_correctness_verifiability}

The packages with higher scores for correctness and verifiability used a wider array of
techniques to improve correctness, and had better documentation to witness this.
For instance, we looked for evidence of unit testing, and found evidence for
only about half of the projects. We identified five projects using continuous
integration tools: \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji},
\textit{dwv}, and \textit{OHIF Viewer}.

Even for projects with well-organized documentation, requirements
specifications and theory manuals were still missing.
The only requirements-related document we found was a road
map of \textit{3D Slicer}, which contained design requirements for upcoming
changes.

\subsection{Surface Reliability} \label{sec_result_reliability}

We were able to follow the steps in the tutorials that existed (seven packages
had them.) However, \textit{GATE} could not open macro files and became
unresponsive several times, without any descriptive error message. We found that
\textit{Drishti} crashed when loading damaged image files, without showing any
descriptive error message.

\subsection{Surface Robustness} \label{sec_result_robustness}

The packages with higher robustness scores gracefully handled unexpected or unanticipated
inputs, typically showing a clear error message. We may have underscored
\textit{OHIF Viewer}, since we needed further customization to load
data.

According to their documentation, all 29 software packages should support the
DICOM standard. To test robustness, we prepared two types of image files:
correct and incorrect formats (with the incorrect format created by relabelling a
text file to have the ``.dcm'' extension).  All software packages loaded the
correct format image, except for \textit{GATE}, which failed for unknown
reasons.  For the broken format, \textit{MatrixUser}, \textit{dwv}, and
\textit{Slice:Drop} ignored the incorrect format,
did not show any error message and displayed a blank image.
\textit{MRIcroGL} behaved similarly except that it showed a meaningless image.
\textit{Drishti} successfully detected the broken format of the file, but the
software crashed as a result.

\subsection{Surface Usability} \label{sec_result_usability}

The software with higher usability scores usually provided both comprehensive
documented guidance and a good user experience. \textit{INVESALIUS 3} provided
an excellent example of a detailed and precise user manual. \textit{GATE} also
provided numerous documents, but unfortunately we had difficulty understanding
and using them. We found getting started tutorials for only 11 projects, but a
user manual for 22 projects. \textit{MRIcroGL} was the only project that
explicitly documented expected user characteristics.

\subsection{Maintainability} \label{sec_score_maintainability}

We gave \textit{3D Slicer} the highest score for maintainability because we
found it had the most comprehensive artifacts. Only a few of the 29 projects had
a product, developer's manual, or API (Application Programming Interface)
documentation, and only \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}
included all three documents (see Table~\ref{tab_maintainability_docs} for the
full data).  Moreover, \textit{3D Slicer} has a much higher percentage of closed
issues (92\%) compared to \textit{ImageJ} (52\%) and \textit{Fiji} (64\%).

\begin{table*}[ht]
\centering
\begin{tabular}{lccc}
\toprule
\multicolumn{1}{c}{Software} & Prod.\ Roadmap & Dev.\ Manual & API Doc. \\ 
\midrule
3D Slicer & \checkmark & \checkmark & \checkmark \\
ImageJ & \checkmark & \checkmark & \checkmark \\
Weasis &  & \checkmark &  \\
OHIF Viewer &  & \checkmark & \checkmark \\
Fiji & \checkmark & \checkmark & \checkmark \\
ParaView & \checkmark &  &  \\
SMILI &  &  & \checkmark \\
medInria &  & \checkmark &  \\
INVESALIUS 3 & \checkmark &  &  \\
dwv &  &  & \checkmark \\
BioImage Suite Web &  & \checkmark &  \\
Gwyddion &  & \checkmark & \checkmark \\ 
\bottomrule
\end{tabular}
\caption{Software with the maintainability documents (listed in descending order of 
maintainability score)}
\label{tab_maintainability_docs}
\end{table*}

Twenty-seven of the 29 projects used git for version control, with 24 of
these using GitHub. \textit{AMIDE} used Mercurial and \textit{Gwyddion} used
Subversion. \textit{XMedCon}, \textit{AMIDE}, and \textit{Gwyddion} used
SourceForge. \textit{DicomBrowser} and \textit{3DimViewer} used BitBucket.

\subsection{Reusability} \label{sec_result_reusability}

We have assumed that smaller code files are
likely more reusable -- see Table \ref{tab_loc_per_file} for the details.

\begin{table*}[ht]
\centering
\begin{tabular}{lllll}
\toprule
\multirow{2}{*}{Software} & \multirow{2}{*}{Text Files} & \multirow{2}{*}{Total Lines} & \multirow{2}{*}{LOC} & \multirow{2}{*}{LOC/file} \\
 &  &  &  &  \\ 
\midrule
OHIF Viewer & 1162 & 86306 & 63951 & 55 \\
3D Slicer & 3386 & 709143 & 501451 & 148 \\
Gwyddion & 2060 & 787966 & 643427 & 312 \\
ParaView & 5556 & 1276863 & 886326 & 160 \\
OsiriX Lite & 2270 & 873025 & 544304 & 240 \\
Horos & 2346 & 912496 & 561617 & 239 \\
medInria & 1678 & 214607 & 148924 & 89 \\
Weasis & 1027 & 156551 & 123272 & 120 \\
BioImage Suite Web & 931 & 203810 & 139699 & 150 \\
GATE & 1720 & 311703 & 207122 & 120 \\
Ginkgo CADx & 974 & 361207 & 257144 & 264 \\
SMILI & 275 & 90146 & 62626 & 228 \\
Fiji & 136 & 13764 & 10833 & 80 \\
Drishti & 757 & 345225 & 268168 & 354 \\
ITK-SNAP & 677 & 139880 & 88530 & 131 \\
3DimViewer & 730 & 240627 & 178065 & 244 \\
DICOM Viewer & 302 & 34701 & 30761 & 102 \\
ImageJ & 40 & 10740 & 9681 & 242 \\
dwv & 188 & 71099 & 47815 & 254 \\
MatrixUser & 216 & 31336 & 23121 & 107 \\
INVESALIUS 3 & 156 & 59328 & 48605 & 312 \\
AMIDE & 183 & 139658 & 102827 & 562 \\
Papaya & 110 & 95594 & 71831 & 653 \\
MicroView & 137 & 36173 & 27470 & 201 \\
XMedCon & 202 & 129991 & 96767 & 479 \\
MRIcroGL & 97 & 50445 & 8493 & 88 \\
Slice:Drop & 77 & 25720 & 19020 & 247 \\
DicomBrowser & 54 & 7375 & 5505 & 102 \\
dicompyler & 48 & 19201 & 15941 & 332 \\ 
\bottomrule
\end{tabular}
\caption{Number of files and lines (by reusability scores)}
\label{tab_loc_per_file}
\end{table*}

\subsection{Surface Understandability} \label{sec_result_understandability}

All projects had a consistent coding style with parameters in
the same order for all functions, modularized code, and, clear comments that
indicate what is done, not how. However, we only found explicit identification
of a coding standard for 3 out of the 29: \textit{3D Slicer}, \textit{Weasis},
and \textit{ImageJ}. We also found hard-coded constants (rather than symbolic
constants) in \textit{medInria}, \textit{dicompyler}, \textit{MicroView}, and
\textit{Papaya}. We did not find any reference to the algorithms used in
projects \textit{XMedCon}, \textit{DicomBrowser}, \textit{3DimViewer},
\textit{BioImage Suite Web}, \textit{Slice:Drop}, \textit{MatrixUser},
\textit{DICOM Viewer}, \textit{dicompyler}, and \textit{Papaya}.

\subsection{Visibility/Transparency} \label{sec_result_visibility_transparency}

Generally speaking, the teams that actively documented
their development process and plans scored higher.
Table~\ref{tab_Visibility/Transparency_docs} shows the projects that had
documents for the development process, project status, development environment,
and release notes.

\begin{table*}[!ht]
\centering
\begin{tabular}{lllll}
\toprule
Software & Dev.\ Process & Proj.\ Status & Dev.\ Env. & Rls.\ Notes \\ 
\midrule
3D Slicer & \checkmark & \checkmark & \checkmark & \checkmark \\
ImageJ & \checkmark & \checkmark & \checkmark & \checkmark \\
Fiji & \checkmark & \checkmark & \checkmark &  \\
MRIcroGL &  &  &  & \checkmark \\
Weasis &  &  & \checkmark & \checkmark \\
ParaView &  & \checkmark &  &  \\
OHIF Viewer &  &  & \checkmark & \checkmark \\
DICOM Viewer &  &  & \checkmark & \checkmark \\
medInria &  &  & \checkmark & \checkmark \\
SMILI &  &  &  & \checkmark \\
Drishti &  &  &  & \checkmark \\
INVESALIUS 3 &  &  &  & \checkmark \\
OsiriX Lite &  &  &  & \checkmark \\
GATE &  &  &  & \checkmark \\
MicroView &  &  &  & \checkmark \\
MatrixUser &  &  &  & \checkmark \\
BioImage Suite Web &  &  & \checkmark &  \\
ITK-SNAP &  &  &  & \checkmark \\
Horos &  &  &  & \checkmark \\
dwv &  &  &  & \checkmark \\
Gwyddion &  &  &  & \checkmark \\ 
\bottomrule
\end{tabular}
\caption{Software with visibility/transparency related documents (listed in
descending order of visibility/transparency score)}
\label{tab_Visibility/Transparency_docs}
\end{table*}

\section{Discussion}

We first compare our ranking to a (proxy for) the community's ranking. We then
compare the state of the practice for MI with that of other research software.
In particular we provide details on recommended artifacts that are rarely observed
for MI software. [other new sections] Section~\ref{sec_threats_to_validity} presents threats to the
validity of our data and conclusions.

\subsection{Comparison to Community Ranking} \label{Sec_VsCommunityRanking}

We use GitHub stars, number of forks and number of people watching the projects
are proxies for community ranking -- see Table~\ref{tab_ranking_vs_GitHub}
for statistics collected in July 2021.  Recall that 24 projects use GitHub.
Our ranking and GitHub popularity, at least for the top five projects,
seems to line up fairly well.

We ranked some popular packages fairly low, such as \textit{dwv}. This is
because we were unable to build it locally, even though we followed its
installation instructions. However, we were able to use its web
version for the rest of the measurements. Additionally, this version 
did not detect a broken DICOM file and instead displayed a blank image
(Section~\ref{sec_result_robustness}).
\textit{DICOM Viewer} ranked low as we were unable to install the
NextCloud platform.

@@

\subsection{Software Artifacts}
\label{Sec_CompareArtifacts}

We use nine research software development guidelines to compare recommended 
software artifacts versus those present in MI software.
These guidelines are:
\begin{itemize}
\item United States Geological Survey Software Planning Checklist
\cite{USGS2019},
\item DLR (German Aerospace Centre) Software Engineering Guidelines
\cite{TobiasEtAl2018}, 
\item Scottish Covid-19 Response Consortium Software Checklist
\cite{BrettEtAl2021},
\item Good Enough Practices in Scientific Computing \cite{WilsonEtAl2016},
\item xSDK (Extreme-scale Scientific Software Development Kit) Community Package
Policies \cite{SmithAndRoscoe2018},
\item Trilinos Developers Guide \cite{HerouxEtAl2008},
\item EURISE (European Research Infrastructure Software Engineers') Network
Technical Reference \cite{ThielEtAl2020},
\item CLARIAH (Common Lab Research Infrastructure for the Arts and Humanities)
Guidelines for Software Quality \cite{vanGompelEtAl2016}, and
\item A Set of Common Software Quality Assurance Baseline Criteria for Research
Projects \cite{OrvizEtAl2017}.
\end{itemize}

\begin{table*}[ht]
%\begin{center}
\begin{tabular}{ p{2.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm} }
\toprule
~ \ & \cite{USGS2019} & \cite{TobiasEtAl2018} & \cite{BrettEtAl2021} &
\cite{WilsonEtAl2016} & \cite{SmithAndRoscoe2018} & \cite{HerouxEtAl2008} &
\cite{ThielEtAl2020} & \cite{vanGompelEtAl2016} & \cite{OrvizEtAl2017} & MI\\
\midrule
LICENSE & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & &
\checkmark & \checkmark & \checkmark & C\\
README &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark &
\checkmark & \checkmark & C\\
CONTRIBUTING &  & \checkmark & \checkmark & \checkmark & \checkmark & &
\checkmark & \checkmark & \checkmark & R\\
CITATION &  &  &  & \checkmark & & & & \checkmark & \checkmark & U\\
CHANGELOG &  & \checkmark &  & \checkmark & \checkmark & & \checkmark &  &  & U\\
INSTALL &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & U\\
\midrule
Uninstall &  &  &  &  & & & & \checkmark & &  \\
Dependency List &  &  & \checkmark & & \checkmark & & & \checkmark &  & R\\
Authors &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & U\\
Code of Conduct &  &  &  &  & & & \checkmark & & & R\\
Acknowledgements &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & U\\
Code Style Guide &  & \checkmark &  &  & & & \checkmark & \checkmark & \checkmark & R\\
Release Info. &  & \checkmark &  &  & & \checkmark & \checkmark & & & C\\
Prod.\ Roadmap &  &  &  &  & & \checkmark & \checkmark & \checkmark & & R\\
\midrule
Getting started &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
User manual &  &  & \checkmark &  & & & \checkmark & & & C\\
Tutorials &  &  &  &  & & & \checkmark & & & U\\
FAQ &  &  &  &  & & & \checkmark & \checkmark & \checkmark & U\\
\midrule
Issue Track &  & \checkmark & \checkmark & & \checkmark & \checkmark &
\checkmark & & \checkmark & C\\
Version Control &  & \checkmark & \checkmark & \checkmark & \checkmark &
\checkmark & \checkmark & \checkmark & \checkmark & C\\ 
Build Scripts &  & \checkmark &  & \checkmark & \checkmark & \checkmark &
\checkmark & & \checkmark & U\\
\midrule
Requirements &  & \checkmark &  &  & & \checkmark &  &  & \checkmark & R\\
Design Doc.\ &  & \checkmark  & \checkmark &  & \checkmark & & \checkmark &
\checkmark& \checkmark & R\\
API Doc. &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
Test Plan &  & \checkmark &  &  & & \checkmark & & & &  \\
Test Cases & \checkmark & \checkmark & \checkmark &  & \checkmark & \checkmark &
\checkmark & \checkmark & \checkmark & U\\
\bottomrule
\end{tabular}
\caption{Comparison of Recommended Artifacts in Software Development Guidelines
to Artifacts in MI Projects (C for Common, U for Uncommon and R for Rare)}
\label{Tbl_Guidelines}
%\end{center}
\end{table*}

@@

\subsection{Interview Methods} \label{sec_interview_methods}

The repository-based measurements summarize the information we can collect from
on-line resources. This information is incomplete because it doesn't generally
capture the development process, the developer pain points, the perceived
threats to software quality, and the developers' strategies to address these
threats.  Therefore, part of our methodology involves interviewing developers.

>>>>

Our investigation is based on the following four research questions:
\begin{enumerate}
\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
{open source} software projects exist? (Section~\ref{SecWhatProjects})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Based on
quantitative measurements of each project's development practices, which
projects follow best practices? (Section~\ref{SecMeasurementResults})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
similar are the top projects identified in \rqref{RQ_HighestQuality}
to the most popular projects as viewed by the community?
(Section~\ref{Sec_VsCommunityRanking})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
do MI projects compare to general research software with respect to the
artifacts (documents, scripts and code) present in their repositories?
(Section~\ref{Sec_CompareArtifacts})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:] How
do MI projects compare to research software in general with respect to the use
of tools (Section~\ref{Sec_CompareTools}) for development
(\rqref{RQ_CompareToolsProjMngmnt}.a); and, project management
(\rqref{RQ_CompareToolsProjMngmnt}.b)?
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
How do MI projects compare to research software in general with respect to
principles, processes, and methodologies used?
(Section~\ref{Sec_CompareMethodologies})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
the pain points for developers working on MI software projects?
(Section~\ref{painpoints})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
do the pain points of developers from MI compare to the pain points
for research software in general? (Section~\ref{painpoints})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For MI
developers what specific best practices are taken to address the pain points
and software quality concerns? (Section~\ref{painpoints})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
What research software development practice could potentially address the
pain point concerns identified in \rqref{RQ_PainPoints}?
(Section~\ref{ch_recommendations})
\end{enumerate}

@@

\subsubsection{Interview Methods} \label{sec_interview_methods}

The repository-based measurements are incomplete because they don't generally
capture the development process, the developer pain points, the perceived
threats to software quality, and the developers' strategies to address these
threats.  Therefore, part of our methodology involves interviewing developers.
We based our interviews on a list of 20 questions, which can be found in
[citation redacted for double blind].
%Smith et al. \cite{SmithEtAl2021}. 
Some questions are about the background of the software, the development teams,
the interviewees, and how they organize their projects.  We also ask about the
developer's understanding of the users. Some questions focus on the current and
past difficulties, and the solutions the team has found, or plan to try. We also
discuss documentation, both with respect to how it is currently done, and how it
is perceived. A few questions are about specific software qualities, such as
maintainability, understandability, usability, and reproducibility. The
interviews are semi-structured based on the question list.

\section{In-Scope Open-Source MI Software} \label{SecWhatProjects}

@@

\section{Which Projects Follow Best Practices?} \label{SecMeasurementResults}

@@

The overall score is based on the measurement of the identified qualities. Full
details of how the projects compare for each quality can be found in [citation
redacted for blind review].
%% DOUBLE BLIND
%Smith et al.~\cite{SmithEtAl2024_MI_SOP}
The highlights are as follows:

\begin{description}

\item [Installability] We found installation instructions for 16 projects, but
two did not need them (\textit{BioImage Suite Web} and \textit{Slice:Drop}) as
they are web applications. 10 of the projects required extra dependencies: Five
depend on a specific browser; \textit{dwv}, \textit{OHIF Viewer}, and
\textit{GATE} needs extra libraries to build; \textit{ImageJ} and \textit{Fiji}
need an unzip tool; \textit{MatrixUser} needs Matlab; \textit{DICOM Viewer}
needs a Nextcloud platform. \textit{GATE}, \textit{dwv}, and \textit{DICOM
Viewer} showed severe installation problems. We were not able to install them,
even after a reasonable amount of time (2 hours).

\item [Correctness \& Verifiability] The packages with higher scores for
correctness and verifiability used a wider array of techniques to improve
correctness, and had better documentation to witness this.  For instance, we
looked for evidence of unit testing, and found evidence for only about half of
the projects. We identified five projects using continuous integration tools:
\textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and
\textit{OHIF Viewer}. The only requirements-related document we found was a road
map of \textit{3D Slicer}, which contained design requirements for upcoming
changes.

\item [Surface Reliability] We were able to follow the steps in the tutorials
that existed (seven packages had them.) However, \textit{GATE} could not open
macro files and became unresponsive several times, without any descriptive error
message. We found that \textit{Drishti} crashed when loading damaged image
files, without showing any descriptive error message.

\item [Surface Robustness] According to their documentation, all 29 software
packages should support the DICOM standard. To test robustness, we prepared two
types of image files: correct and incorrect formats (with the incorrect format
created by relabelling a text file to have the ``.dcm'' extension).  All
software packages loaded the correct format image, except for \textit{GATE},
which failed for unknown reasons.  For the broken format, \textit{MatrixUser},
\textit{dwv}, and \textit{Slice:Drop} ignored the incorrect format, did not show
any error message and displayed a blank image.  \textit{MRIcroGL} behaved
similarly except that it showed a meaningless image.  \textit{Drishti}
successfully detected the broken format of the file, but the software crashed as
a result.

\item [Surface Usability] The software with higher usability scores usually
provided both comprehensive documented guidance and a good user experience.
\textit{INVESALIUS 3} provided an excellent example of a detailed and precise
user manual. \textit{GATE} also provided numerous documents, but unfortunately
we had difficulty understanding and using them. We found getting started
tutorials for only 11 projects, but a user manual for 22 projects.
\textit{MRIcroGL} was the only project that explicitly documented expected user
characteristics.

\item [Maintainability] We gave \textit{3D Slicer} the highest score for
maintainability because we found it had the most comprehensive artifacts. Only a
few of the 29 projects had a product, developer's manual, or API (Application
Programming Interface) documentation, and only \textit{3D Slicer},
\textit{ImageJ}, \textit{Fiji} included all three documents.  Moreover,
\textit{3D Slicer} has a much higher percentage of closed issues (92\%) compared
to \textit{ImageJ} (52\%) and \textit{Fiji} (64\%). Twenty-seven of the 29
projects used git for version control, with 24 of these using GitHub.
\textit{AMIDE} used Mercurial and \textit{Gwyddion} used Subversion.
\textit{XMedCon}, \textit{AMIDE}, and \textit{Gwyddion} used SourceForge.
\textit{DicomBrowser} and \textit{3DimViewer} used BitBucket.

\item [Reusability] We have assumed that smaller code files are
likely more reusable -- see Table \ref{tab_loc_per_file} for the details.

\item [Surface Understandability] All projects had a consistent coding style
with parameters in the same order for all functions, modularized code, and,
clear comments that indicate what is done, not how. However, we only found
explicit identification of a coding standard for 3 out of the 29: \textit{3D
Slicer}, \textit{Weasis}, and \textit{ImageJ}. We also found hard-coded
constants (rather than symbolic constants) in \textit{medInria},
\textit{dicompyler}, \textit{MicroView}, and \textit{Papaya}. We did not find
any reference to the algorithms used in projects \textit{XMedCon},
\textit{DicomBrowser}, \textit{3DimViewer}, \textit{BioImage Suite Web},
\textit{Slice:Drop}, \textit{MatrixUser}, \textit{DICOM Viewer},
\textit{dicompyler}, and \textit{Papaya}.

\item [Visibility/Transparency] Generally speaking, the teams that actively
documented their development process and plans scored higher.  \textit{3D
Slicer} and \textit{ImageJ} were the only projects to include documentation for
all of the following: development process, project status, development
environment and release notes.

\end{description}

\begin{table*}[!ht]
\centering
\begin{tabular}{lllll}
\toprule
\multirow{2}{*}{Software} & \multirow{2}{*}{Text Files} & \multirow{2}{*}{Total Lines} & \multirow{2}{*}{LOC} & \multirow{2}{*}{LOC/file} \\
 &  &  &  &  \\ 
\midrule
OHIF Viewer & 1162 & 86306 & 63951 & 55 \\
3D Slicer & 3386 & 709143 & 501451 & 148 \\
Gwyddion & 2060 & 787966 & 643427 & 312 \\
ParaView & 5556 & 1276863 & 886326 & 160 \\
OsiriX Lite & 2270 & 873025 & 544304 & 240 \\
Horos & 2346 & 912496 & 561617 & 239 \\
medInria & 1678 & 214607 & 148924 & 89 \\
Weasis & 1027 & 156551 & 123272 & 120 \\
BioImage Suite Web & 931 & 203810 & 139699 & 150 \\
GATE & 1720 & 311703 & 207122 & 120 \\
Ginkgo CADx & 974 & 361207 & 257144 & 264 \\
SMILI & 275 & 90146 & 62626 & 228 \\
Fiji & 136 & 13764 & 10833 & 80 \\
Drishti & 757 & 345225 & 268168 & 354 \\
ITK-SNAP & 677 & 139880 & 88530 & 131 \\
3DimViewer & 730 & 240627 & 178065 & 244 \\
DICOM Viewer & 302 & 34701 & 30761 & 102 \\
ImageJ & 40 & 10740 & 9681 & 242 \\
dwv & 188 & 71099 & 47815 & 254 \\
MatrixUser & 216 & 31336 & 23121 & 107 \\
INVESALIUS 3 & 156 & 59328 & 48605 & 312 \\
AMIDE & 183 & 139658 & 102827 & 562 \\
Papaya & 110 & 95594 & 71831 & 653 \\
MicroView & 137 & 36173 & 27470 & 201 \\
XMedCon & 202 & 129991 & 96767 & 479 \\
MRIcroGL & 97 & 50445 & 8493 & 88 \\
Slice:Drop & 77 & 25720 & 19020 & 247 \\
DicomBrowser & 54 & 7375 & 5505 & 102 \\
dicompyler & 48 & 19201 & 15941 & 332 \\ 
\bottomrule
\end{tabular}
\caption{Number of files and lines (by reusability scores)}
\label{tab_loc_per_file}
\end{table*}

\section{Comparison to Community Ranking} \label{Sec_VsCommunityRanking}

To answer \rqref{RQ_CompareHQ2Popular} we use GitHub stars, number of forks and
number of people watching the projects are proxies for community ranking.
Table~\ref{tab_ranking_vs_GitHub} show the statistics for data collected in July
2021.  Recall that 24 projects use GitHub (Section~\ref{SecMeasurementResults}).

Our ranking and GitHub popularity, at least for the top five projects, seems to
line up fairly well. However, we ranked some popular packages fairly low, such
as \textit{dwv}. This is because we were unable to build it locally, even though
we followed its installation instructions. However, we were able to use its web
version for the rest of the measurements. Additionally, this version did not
detect a broken DICOM file and instead displayed a blank image
(Section~\ref{SecMeasurementResults}).  \textit{DICOM Viewer} ranked low as we
were unable to install the NextCloud platform.

@@

\section{Compare to Recommended Software Artifacts} \label{Sec_CompareArtifacts}

For \rqref{RQ_CompareArtifacts} we use nine research software development
guidelines to compare recommended software artifacts versus those present in MI
software.  These guidelines are:
\begin{itemize}
\item United States Geological Survey Software Planning Checklist
\cite{USGS2019},
\item DLR (German Aerospace Centre) Software Engineering Guidelines
\cite{TobiasEtAl2018}, 
\item Scottish Covid-19 Response Consortium Software Checklist
\cite{BrettEtAl2021},
\item Good Enough Practices in Scientific Computing \cite{WilsonEtAl2016},
\item xSDK (Extreme-scale Scientific Software Development Kit) Community Package
Policies \cite{SmithAndRoscoe2018},
\item Trilinos Developers Guide \cite{HerouxEtAl2008},
\item EURISE (European Research Infrastructure Software Engineers') Network
Technical Reference \cite{ThielEtAl2020},
\item CLARIAH (Common Lab Research Infrastructure for the Arts and Humanities)
Guidelines for Software Quality \cite{vanGompelEtAl2016}, and
\item A Set of Common Software Quality Assurance Baseline Criteria for Research
Projects \cite{OrvizEtAl2017}.
\end{itemize}

\begin{table*}[!ht]
\begin{center}
\begin{tabular}{ p{3.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm} }
\toprule
~ \ & \cite{USGS2019} & \cite{TobiasEtAl2018} & \cite{BrettEtAl2021} &
\cite{WilsonEtAl2016} & \cite{SmithAndRoscoe2018} & \cite{HerouxEtAl2008} &
\cite{ThielEtAl2020} & \cite{vanGompelEtAl2016} & \cite{OrvizEtAl2017} & MI\\
\midrule
LICENSE & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & &
\checkmark & \checkmark & \checkmark & C\\
README &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark &
\checkmark & \checkmark & C\\
CONTRIBUTING &  & \checkmark & \checkmark & \checkmark & \checkmark & &
\checkmark & \checkmark & \checkmark & R\\
CITATION &  &  &  & \checkmark & & & & \checkmark & \checkmark & U\\
CHANGELOG &  & \checkmark &  & \checkmark & \checkmark & & \checkmark &  &  & U\\
INSTALL &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & U\\
\midrule
Uninstall &  &  &  &  & & & & \checkmark & &  \\
Dependency List &  &  & \checkmark & & \checkmark & & & \checkmark &  & R\\
Authors &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & U\\
Code of Conduct &  &  &  &  & & & \checkmark & & & R\\
Acknowledgements &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & U\\
Code Style Guide &  & \checkmark &  &  & & & \checkmark & \checkmark & \checkmark & R\\
Release Info. &  & \checkmark &  &  & & \checkmark & \checkmark & & & C\\
Prod.\ Roadmap &  &  &  &  & & \checkmark & \checkmark & \checkmark & & R\\
\midrule
Getting started &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
User manual &  &  & \checkmark &  & & & \checkmark & & & C\\
Tutorials &  &  &  &  & & & \checkmark & & & U\\
FAQ &  &  &  &  & & & \checkmark & \checkmark & \checkmark & U\\
\midrule
Issue Track &  & \checkmark & \checkmark & & \checkmark & \checkmark &
\checkmark & & \checkmark & C\\
Version Control &  & \checkmark & \checkmark & \checkmark & \checkmark &
\checkmark & \checkmark & \checkmark & \checkmark & C\\ 
Build Scripts &  & \checkmark &  & \checkmark & \checkmark & \checkmark &
\checkmark & & \checkmark & U\\
\midrule
Requirements &  & \checkmark &  &  & & \checkmark &  &  & \checkmark & R\\
Design Doc.\ &  & \checkmark  & \checkmark &  & \checkmark & & \checkmark &
\checkmark& \checkmark & R\\
API Doc. &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
Test Plan &  & \checkmark &  &  & & \checkmark & & & &  \\
Test Cases & \checkmark & \checkmark & \checkmark &  & \checkmark & \checkmark &
\checkmark & \checkmark & \checkmark & U\\
\bottomrule
\end{tabular}
\caption{Comparison of Recommended Artifacts in Software Development Guidelines
to Artifacts in MI Projects (C for Common, U for Uncommon and R for Rare)}
\label{Tbl_Guidelines}
\end{center}
\end{table*}

@@
